### 2014-09-11 text/Chapter04.tex
Examples of structural comonads (Chapter 5)
When is coeffect a monad (Chapter 4)

\begin{example}[Indexed option]
Flat coeffect algebra for liveness defines $\cpar$ and $\czip$ as $\sqcup$ and $\sqcap$, respectively 
and specifies that $\ident{D} \sqsubseteq \ident{L}$. Recall also that the object mapping is defined 
as $\ctyp{\ident{L}}{\alpha} = \alpha$ and $\ctyp{\ident{D}}{\alpha} = 1$. The additional operations 
of a flat indexed comonad are defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{merge}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
\ident{merge}_{\ident{L}, \ident{D}}~(a, ()) &\narrow{=}& ()\\
\ident{merge}_{\ident{D}, \ident{L}}~((), b) &\narrow{=}& ()\\
\ident{merge}_{\ident{D}, \ident{D}}~((), ()) &\narrow{=}& ()\\
\end{array}
\quad
\begin{array}{rcl}
\ident{split}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
\ident{split}_{\ident{L}, \ident{D}}~(a, b) &\narrow{=}& (a, ())\\
\ident{split}_{\ident{D}, \ident{L}}~(a, b) &\narrow{=}& ((), b)\\
\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& (a, b)\\
\end{array}
\quad
\begin{array}{rcl}
\ident{lift}_{\ident{L}, \ident{D}}~v &\narrow{=}& ()\\
\ident{lift}_{\ident{L}, \ident{L}}~v &\narrow{=}& v\\
\ident{lift}_{\ident{D}, \ident{D}}~() &\narrow{=}& ()\\
\end{array}
\end{equation*}
\end{example}

@@

\begin{equation*}
\begin{array}{l}
\ident{counit}_{0}\langle a_0 \rangle = a_0
\\[0.45em]
\ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle = \\[-0.25em]
\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
\\[0.45em]
\ident{merge}_{m, n} (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle) = \\[-0.25em]
\quad \langle (a_0, b_0), \ldots, (a_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle
\\[0.45em]
\ident{split}_{m, n} \langle (a_0, b_0), \ldots, (a_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle = \\[-0.25em]
\quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
\\[0.45em]
\ident{lift}_{n', n} \langle a_0, \ldots, a_n' \rangle = \qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
\quad \langle a_0, \ldots, a_n \rangle
\end{array}
\begin{array}{l}
\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[11em]
~\\
\end{array}
\end{equation*}
\end{example}

@@

\newcommand{\ctx}{\textit{ctx}}
\begin{figure*}[t]

@@

\section{Related work}
\label{sec:flat-related-monads}

Most of the related work has been already discussed in Chapter~\ref{ch:pathways} or throughout
the chapter. In this section, we look in more details at the relationship between coeffects,
comonads and monads.

\subsection{When coeffect is not a monad}
Coeffect systems differ from effect systems in three important ways:

@@

\subsection{When coeffect is a monad}
As discussed in \ref{sec:applications-flat-impl} ...

Product comonad

$C_{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma)$

Reader monad

$M_{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau$

\begin{equation*}
\begin{array}{l}
 \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 = \\
 (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
 \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) \\
 \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \\
\end{array}
\end{equation*}

The difference between effects and coeffects is mainly in the (\emph{abs}) rule. While the 
semantic model (monads vs. comonads) is very different, we can consider extending the two to 
obtain equivalent syntactic rules. To allow splitting of implicit parameters in lambda abstraction, 
the reader monad needs an operation that eagerly performs some effects of a function:

\begin{equation*}
(\tau_1 \rightarrow \mtyp{\cclrd{r} \cseq \cclrd{s}}{\tau_2}) \rightarrow 
  \mtyp{\cclrd{r}}{(\tau_1 \rightarrow \mtyp{\cclrd{s}}{\tau_2})}
\end{equation*}

To obtain a pure lambda abstraction
for coeffects, we need to restrict the $\cpar_{r, s}$ 
operation of indexed comonads, so that the first parameter is annotated with $\cunit$ (meaning
no effects): $\C^{\cunit} A \times \C^r B \rightarrow \C^{r}(A \times B)$.

\begin{equation*}
\begin{array}{rcl}
 \sem{\lambda x.e} &=& \ident{unit}~(\lambda x. \sem{e}) \\
 \sem{\lambda x.e} &=& \ident{unit}~(\lambda x. \sem{e}) \\
\end{array}
\end{equation*}
% ==================================================================================================

\section{Conclusions}

We examined three simple calculi with associated
static analyses (liveness analysis, implicit parameters, and dataflow
analysis). These were unified in the \emph{coeffect calculus},
providing a general coeffect system parameterised by an
algebraic structure describing the propagation of context
requirements throughout a program.

We model the semantics of coeffect calculus using \emph{indexed comonad} -- a novel structure, which
is more powerful than (monoidal) comonads. Indices of the indexed comonad operations manifest the 
semantic propagation of context such that the propagation of information in the general coeffect
type system corresponds exactly to the semantic propagation of context in our categorical model.

We consider the analysis of context to be essential, not least for the examples here but 
also given increasingly rich and diverse distributed systems.

>>>>

\begin{example}[Indexed option]
Flat coeffect algebra for liveness defines $\cpar$ and $\czip$ as $\sqcup$ and $\sqcap$, respectively 
and specifies that $\ident{D} \sqsubseteq \ident{L}$. Recall also that the object mapping is defined 
as $\ctyp{\ident{L}}{\alpha} = \alpha$ and $\ctyp{\ident{D}}{\alpha} = 1$. The additional operations 
of a flat indexed comonad are defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{merge}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
\ident{merge}_{\ident{L}, \ident{D}}~(a, ()) &\narrow{=}& ()\\
\ident{merge}_{\ident{D}, \ident{L}}~((), b) &\narrow{=}& ()\\
\ident{merge}_{\ident{D}, \ident{D}}~((), ()) &\narrow{=}& ()\\
\end{array}
\quad
\begin{array}{rcl}
\ident{split}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
\ident{split}_{\ident{L}, \ident{D}}~(a, b) &\narrow{=}& (a, ())\\
\ident{split}_{\ident{D}, \ident{L}}~(a, b) &\narrow{=}& ((), b)\\
\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& ((), ()))\\
\end{array}
\quad
\begin{array}{rcl}
\ident{lift}_{\ident{L}, \ident{D}}~v &\narrow{=}& ()\\
\ident{lift}_{\ident{L}, \ident{L}}~v &\narrow{=}& v\\
\ident{lift}_{\ident{D}, \ident{D}}~() &\narrow{=}& ()\\
\end{array}
\end{equation*}
\end{example}

@@

\begin{equation*}
\begin{array}{l}
\ident{counit}_{0}\langle a_0 \rangle = a_0
\\[0.45em]
\ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle = \\[-0.25em]
\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots, f \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
\\[0.45em]
\ident{merge}_{m, n} (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle) = \\[-0.25em]
\quad \langle (a_0, b_0), \ldots, (a_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle
\\[0.45em]
\ident{split}_{m, n} \langle (a_0, b_0), \ldots, (a_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle = \\[-0.25em]
\quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
\\[0.45em]
\ident{lift}_{n', n} \langle a_0, \ldots, a_{n'} \rangle = \qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
\quad \langle a_0, \ldots, a_n \rangle
\end{array}
\begin{array}{l}
\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[11em]
~\\
\end{array}
\end{equation*}
\end{example}

@@

\begin{figure*}[t]

@@

\section{Related work}
\label{sec:flat-related}

Most of the related work leading to coeffects has already been discussed in Chapter~\ref{ch:pathways} 
and we covered work related to individual concepts throughout the chapter. In this section, we
do not repeat the discussion present elsewhere -- we discuss one specific question that often 
arises when discussing coeffects and that is \emph{when is coeffect (not) an effect?}

We start with a quick overview of the ways in which effects and coeffects differ and then
we briefly look at one (but illustrative) example where the two concepts overlap. We focus 
mainly on the equivalence between the \emph{categorical semantics}, which reveals the nature
of the computations -- rather than considering just the syntactic aspects of the type system.

\subsection{When is coeffect not a monad}
Coeffect systems differ from effect systems in three important ways:

@@

\subsection{When is coeffect a monad}
\label{sec:flat-related-monads}

As discussed in Section~\ref{sec:applications-flat-impl}, one of our examples, implicit 
parameters, can be also captured by a monad. However, \emph{just} a monad is not enough
because lambda abstraction in effect systems does not provide a way of splitting the context
requirements between declaration-site and call-site (or combining the implicit parameters
available in the scope where the function is deined and those specified by the caller).

\paragraph{Categorical relationship.}
Before looking at the necessary extensions, consider the two ways of modelling implicit 
parameters. We assume that the function $\cclrd{r} \rightarrow \sigma$ is a lookup function
for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
are:
%
\begin{equation*}
\begin{array}{lll}
 \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
 \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
\end{array} 
\end{equation*}
%
The \emph{product comonad} simply pairs the value $\tau$ with the lookup function, while
the \emph{reader monad} is a function that, given a lookup function, produces a $\tau$ value.
As noted by Orchard \cite{comonads-vs-monads}, when used to model computation semantics, the 
two representations are equivalent:
%
\begin{remark}
Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using
the currying/uncurrying isomoprhism.
\end{remark}
\begin{proof}
The isomorphism is demonstrated by the following equation:
\begin{equation*}
\begin{array}{l}
 \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 = \\
 (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
 \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) \\
 \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
\end{array}
\end{equation*}
\end{proof}

\noindent
The equivalence holds for monads and comonads (as well as \emph{indexed} monads
and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging.

\paragraph{Delaying effects in monads.} 
In the syntax of the language, the above difference is manifested in the (\emph{abs}) rules for 
monadic effect systems and comonadic coeffect systems. The following listing shows the two rules 
side-by-side, using the coeffect notation for both of them:
%
\begin{equation*}
\tyrule{cabs}
  { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \vdash e : \tau_2 }
  { \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\quad
\tyrule{mabs}
  { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \vdash e : \tau_2 }
  { \coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \tau_2 }
\qquad
\end{equation*}
%
In the comonadic (\emph{cabs}) rule, the implicit parameters of the body are split. However,
the monadic rule (\emph{mabs}) places all requirements on the call-site. This follows from the
fact that monadic semantics uses the \ident{unit} operation in the interpretation of lambda abstraction:
%
\begin{equation*}
\sem{\lambda x.e} \;=\; \ident{unit}~(\lambda x. \sem{e})
\end{equation*}
%
The type of $\ident{unit}$ is $\alpha\rightarrow\mtyp{\alpha}{\cclrd{\emptyset}}$, but in this specific
case, the $\alpha$ is instantiated to be $\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}$ and so this
use of \ident{unit} has a type:
%
\begin{equation*}
\ident{unit}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{\emptyset}}(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2})
\end{equation*}
%
In order to split the implicit parameters of the body ($\cclrd{r \cup s}$ on the left-hand side) between 
the declaration-site ($\cclrd{\emptyset}$ on the outer $\mtyp{}$ on the right-hand side) and the 
call-site ($\cclrd{r \cup s}$ on the inner $\mtyp{}$ on the right-hand side), we need an operation
(which we call \ident{delay}) with the following signature:
%
\begin{equation*}
\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{r}}(\tau_1\rightarrow\mtyp{\cclrd{s}}{\tau_2})
\end{equation*}
%
The operation reveals the difference between effects and coeffects -- intuitively, given a function
with effects $\cclrd{r \cup s}$, it should execute the effects $\cclrd{r}$ when wrapping the 
function, \emph{before} the function actually perforsm the effectful operation with the effects.
The remaining effects $\cclrd{s}$ are delayed as usual, while effects $\cclrd{r}$ are removed
from the effect annotation of the body.

Another important aspect of the signature is that the function needs to be indexed by the coeffect
annotations $\cclrd{r}, \cclrd{s}$. The indices determine how the input context requirements
$\cclrd{r \cup s}$ are split -- and thus guarantee determinism of the function.

The operation cannot be implemented in a useful way for most standard monads, but the 
reader monad is, indeed, an exception. It is not difficult to see how it can be implemented
when we expand the definitions of $\mtyp{\cclrd{r}}\tau$:
%
\begin{equation*}
\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~
(\tau_1\rightarrow(\cclrd{r \cup s} \rightarrow \sigma) \rightarrow \tau_2) \rightarrow 
((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_1\rightarrow(\cclrd{s} \rightarrow \sigma) \rightarrow \tau_2)
\end{equation*}

\paragraph{Restricting coeffects in comonads.} 
As just demonstrated, we can extend monads so that the reader monad is capable of capturing
the semantics of implicit parameters, including the splitting of implicit parameter requirements
in lambda abstraction. Can we also go the other way round and \emph{restrict} the comonadic
semantics so that all requirements are delayed as in the (\emph{mabs}) rule, thus modelling 
fully dynamically scoped parameters?

This is, indeed, also possible. Recall that the semantics of lambda abstraction in the flat
coeffect calculus is modelled using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. The operation takes
two contexts (wrapped in a comonad $\ctyp{\cclrd{r}}\alpha$), combines their carried values and
additional contextual information (implicit parameters). To obtain the (\emph{mabs}) rule, we 
can restrict the first parameter, which corresponds to the declaration-site context:
%
\begin{equation*}
\begin{array}{rcll}
 \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{r}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{r \cup s}}(\alpha \times \beta)
 &\hspace{4em}(\textit{normal}) \\
 \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{\emptyset}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{s}}(\alpha \times \beta)
 &\hspace{4em}(\textit{restricted})
\end{array} 
\end{equation*}
%
In the (\emph{restricted}) version of the operation, the declaration-site context requires
no implicit parameters and so all implicit parameters have to be satisfied by the call-site.
The semantics using the restricted version corresponds to the (\emph{mabs}) rule shown above.

The idea of restricting the operations of the coeffect calculus semantics could be used more
generally. We could allow any of the coeffect algebra operations $\cseq, \czip, \cpar$ to be
\emph{partial} and thus the restricted (fully dynamically-scoped) version of implicit parameters
could be obtained just by changing the definition of $\czip$. Similarly, we could obtain \eg~a
fully lexically-scoped version of the system. Similar idea has been used for the
semantics of effectful computations by Tate~\cite{effects-producer-semantics}.

% ==================================================================================================
%                                                                             
%        ###                         ##                    #                        
%       #   #                         #                                             
%       #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###  
%       #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  #  #     
%       #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###  
%       #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      # 
%        ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####  
%                                                                             
% ==================================================================================================                                                                             
\section{Conclusions}

This chapter presented the \emph{flat coeffect calculus} -- a unified system for tracking 
contextual properties that are \emph{whole-context}, meaning that they are related to the 
execution environment or the enitre context in which computations are executed.
This is the first of the two \emph{coeffect calculi} developed in this thesis.

The flat coeffect calculus is parameterized by a \emph{flat coeffect algebra} that captures
the structure of the information tracked by the type system. We demonstrated how to 
instantiate the system to capture three specific systems discussed earlier in 
Chapter~\ref{ch:applications}, namely liveness, data-flow and implicit parameters.

Next, we introduced the notion of \emph{flat indexed comonad}, which is a generalization of
comonad, equipped with additional operations needed to provide categorical semantics of the 
flat coeffect calculus. The indices of the flat indexed comonad operations correspond to the
coeffect annotations in the type system and provide a foundation for the design of the 
calculus.

Finally, we discussed the equational theory for flat coeffect calculus. Although each 
concrete instance of flat coeffect calculus models different notion of context, there are
syntactic properties that hold for all flat coeffect systems satisfying certain additional
conditions. In particular, two \emph{subject reduction} theorems prove that the operational
semantics for two classes of flat coeffect calculi (including liveness and implicit parameters) 
can be based on call-by-name reduction.

In the upcoming chapter, we move from \emph{flat} coeffect calculi, tracking whole-context
properties to \emph{structural} coeffect calculi, tracking per-variable information, thus
covering systems from the second half of Chapter~\ref{ch:applications}.
